Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-09-18 manuscript/manuscript.tex
added HMM subsection

In addition to publications on the theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
we also recommend the literature on TICA~\cite{tica,tica3,tica2,kinetic-maps},
transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial},
and variational techniques~\cite{noe-vac,vamp-preprint,gmrq},
as these topics play important roles within the standard MSM workflow.

@@

>>>>

In addition to publications on the theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
we also recommend the literature on TICA~\cite{tica,tica3,tica2,kinetic-maps},
transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint},
and variational techniques~\cite{noe-vac,vamp-preprint,gmrq},
as these topics play important roles within the standard MSM workflow.

@@

\subsection*{Hidden Markov state models}

The estimation of an MSM requires the dynamics between microstates to be Markovian.
However, in case of a poor discretization or short trajectories,
it might be hard or even impossible to meet this requirement (as we will see in Notebook~07).

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint}.
HMMs require the dynamics in some hidden (unobserved) state space to be Markovian.

To estimate an HMM, we need a spectral gap after the $m^\textrm{th}$ eigenvalue;
in practice, a timescale separation $t_m \geq 2t_{m+1}$ is sufficient~\cite{pyemma}.
Then, we can approximate the dynamics in the observed microstates ($\mathbf{P}$) at any lag time $k\tau$ via
\begin{equation}
\mathbf{P}(k\tau) \approx \bm{\Pi}^{-1} \bm{\chi}^\top \tilde{\bm{\Pi}} \tilde{\mathbf{P}}^k(\tau) \, \bm{\chi}.
\end{equation}
Here, the $\bm{\Pi}=\left[ \pi_1,\dots,\pi_n \right]$ is a diagonal matrix of the $n$ microstates' stationary probabilities,
$\tilde{\bm{\Pi}}=\left[ \tilde{\pi}_1,\dots,\tilde{\pi}_m \right]$ is a diagonal matrix of the $m$ hidden states' stationary probabilities,
$\tilde{\mathbf{P}}(\tau)$ is a (small) transition matrix between the $m$ hidden states at lag time $\tau$,
and $\bm{\chi}$ is an $m\times n$-dimensional row stochastic matrix
where each row encodes the probabilities of the $n$ microstates to appear in the corresponding hidden state~\cite{noe-proj-hid-msm}.

An HMM estimation usually yields a small model where each hidden state is considered to be metastable and,
thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully.
Given that an HMM is a generalization of an MSM,
we can compute all physico-chemically relevant quantities available for MSMs, too.
